Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-10-27 mprop/mprop.tex
Lots of derivatives

\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}
\newcommand{\DKL}{D_{\mathrm{KL}}}
\newcommand{\pfull}{p(\mathcal{D}, \Theta, \mathbf{X_u}, \mathbf{u}, \mathbf{r})}
\newcommand{\pfullS}{p(\mathcal{D}, \Theta_s, \mathbf{X_u}, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\posterior}{p(\Theta, \mathbf{u}, \mathbf{r} | \mathcal{D}, \mathbf{X_u})}
\newcommand{\approximation}{q_{\bm\nu}(\Theta, \mathbf{u}, \mathbf{r})}
\newcommand{\approximationS}{q_{\bm\nu}(\Theta_s, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\Eq}{\mathbb{E}_{(\Theta, \mathbf{u}, \mathbf{r}) \sim \approximation}}

\section{Introduction}
%briefly explain the context of the project problem

@@

% TODO: positive definite
As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\begin{equation} \label{full}
  \pfull = p(\mathbf{X_u}) \times p(\Theta | \mathbf{X_u}) \times p(\mathbf{u}
  | \Theta, \mathbf{X_u}) \times p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}) \times p(\mathcal{D} | r).
\end{equation}
Here $p(\mathbf{X_u})$ and $p(\Theta | \mathbf{X_u})$ are freely chosen priors,
\begin{align}
  p(\mathbf{u} | \Theta, \mathbf{X_u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \nonumber \\
                                       &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \nonumber \\
                                       &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right) \label{eq:normal}
\end{align}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, where $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in \eqref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k_\Theta(\mathbf{x}_{\mathbf{r},i}, \mathbf{x}_{\mathbf{u},j})$,
where $\mathbf{x}_{\mathbf{r},i}$ and $\mathbf{x}_{\mathbf{u},j}$ denote feature
vectors for the $i$th state in $\mathcal{S}$ and the $j$th state in
$\mathbf{X_u}$, respectively \cite{DBLP:conf/uai/JinDAS17}.

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\Theta$, inducing rewards $\mathbf{u}$, and the reward function $r$. While the
previous paper that considered this IRL model computed maximum likelihood estimates
for $\Theta$ and $\mathbf{u}$, and made an assumption that $\mathbf{r}$ in
\eqref{eq:r} has zero variance \cite{DBLP:conf/nips/LevinePK11}, we aim to
avoid this assumption and use variational inference to approximate the full
posterior distribution $\posterior$. \emph{Variational inference} (VI) is an
approximation technique for probability densities \cite{blei2017variational}.
Let $\approximation$ be our approximating family of probability distributions
for $\posterior$ with its own hyperparameter vector $\bm\nu$. Then it is up
to VI algorithms to optimise $\bm\nu$ in order to minimise the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are and in this case can be defined as follows
\cite{blei2017variational}:
\[ \DKL(\approximation || \posterior) = \Eq \left[ \log
  \frac{\approximation}{\posterior} \right]. \]
Since KL divergence is typically hard to compute, instead of minimising it, VI
typically tries to maximise the \emph{evidence lower bound} (ELBO) defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\Theta\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}

@@

@@

At this point we are forced to make assumptions about the approximate posterior
in order to arrive at an implementable solution. Ways to relax the assumptions
may be investigated towards the end of the project, if time permits.

First, as is common in the literature for applying VI to GPs
\cite{DBLP:conf/nips/ChengB17,DBLP:conf/uai/HensmanFL13,DBLP:conf/icml/HoangHL15,DBLP:journals/jmlr/Titsias09},
we simply set
\begin{equation}
  q(\mathbf{r} | \Theta, \mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}).
\end{equation}
We can make a similarly justified choice for $q(\mathbf{u} | \Theta)$:
\begin{equation}
  q(\mathbf{u} | \Theta) = q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})
\end{equation}
for some variational parameters $\mathbf{m} \in \mathbb{R}^{m}$ and $\mathbf{S}
\in \mathbb{R}^{m \times m}$
\cite{DBLP:conf/nips/ChengB17,DBLP:journals/jmlr/HensmanDS17,DBLP:conf/aaai/HoangHL17}.

In order to have a reasonable way of calculating/approximating the ELBO (see
Section \ref{vi_algs}), we need to have an approximating distribution for
$\Theta$, but unfortunately all the papers in Table \ref{table:approximations}
either fix it or treat it as a variational parameter. Hence we make our first
assumption without justification from previous literature:
\begin{equation} \label{eq:theta}
  q(\Theta) = q(\zeta) \times q(\bm\lambda).
\end{equation}
As it is reasonable for $\bm\lambda$ to take any value in $\mathbb{R}^d$, we can
approximate it simply as
\begin{equation} \label{eq:lambda}
  q(\bm\lambda) = \mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma)
\end{equation}
for some variational parameters $\bm\mu \in \mathbb{R}^d$ and $\bm\Sigma \in
\mathbb{R}^{d \times d}$. In contrast, as the covariance function must produce
non-negative values, we want to restrict $\zeta$ to positive real numbers. At
the same time, the distribution for $\zeta$ should have a flexible mean (i.e.,
not be tied to zero, like in the exponential distribution) and tails that
converge to zero as the value of the random variable moves away from the mean.
We might want to support some right skew, but the distribution should be close
to symmetric with at least some parameter values. This limits our choice of
distributions quite significantly, and we decide to go with the gamma
distribution as it is fairly flexible and commonly used
\cite{hogg2018introduction}. We then define the probability density function of
the gamma distribution as
\begin{equation} \label{eq:zeta}
  q(\zeta) = \Gamma(\zeta; \alpha, \beta) =
  \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta},
\end{equation}
where $\alpha > 0$ and $\beta > 0$ are two parameters of the distribution, and
$\Gamma(\cdot)$ is the gamma function. This gives us our vector of variational
parameters $\bm\nu = (\mathbf{m}, \mathbf{S}, \bm\mu, \bm\Sigma, \alpha,
\beta)^\intercal$.

\subsection{Evidence Lower Bound}

In this section we derive and simplify the ELBO for this (now fully specified)
model. We begin with a lemma that will be useful during the derivation.

% TODO: this is actually the entropy of the multivariate Gaussian
\begin{lemma} \label{expectation_lemma}
  Let $\mathbf{x} \sim \mathcal{N}(\bm\mu, \bm\Sigma)$ be a random Gaussian
  vector with $n$ elements, and let $f$ be its probability density function.
  Then
  \[ \mathbb{E}[\log f(\mathbf{x})] = -\frac{1}{2}\log |\bm\Sigma| + c, \]
  where $c \in \mathbb{R}$ is a constant w.r.t. $\bm\mu$ and $\bm\Sigma$.
\end{lemma}

\begin{proof}
  Similarly to \eqref{eq:normal}, we can write
  \[ \mathbb{E}[\log f(\mathbf{x})] = \mathbb{E}\left[ -\frac{1}{2}(\mathbf{x} -
      \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{x} - \bm\mu) - \frac{1}{2}\log |\bm\Sigma|
      - \frac{n}{2}\log 2\pi \right]. \]
  Note that the term
  \[ \mathbb{E}\left[ -\frac{n}{2}\log 2\pi \right] = -\frac{n}{2}\log 2\pi \]
  can be absorbed into $c$. Now
  \[ \mathbb{E}[(\mathbf{x} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{x} - \bm\mu)] =
    \Tr(\bm\Sigma^{-1}\mathbf{M}) + \mathbf{c}^\intercal\bm\Sigma^{-1}\mathbf{c}, \]
  where $\mathbf{M} = \Var[\mathbf{x} - \bm\mu]$ and $\mathbf{c} =
  \mathbb{E}[\mathbf{x} - \bm\mu]$ \cite{petersen2008matrix}. But
  \[ \mathbf{c} = \mathbb{E}[\mathbf{x} - \bm\mu] = \mathbb{E}[\mathbf{x}] -
    \bm\mu = \bm\mu - \bm\mu = \mathbf{0}, \]
  so
  \[ \mathbb{E}[(\mathbf{x} - \bm\mu)^\intercal\bm\Sigma^{-1}(\mathbf{x} - \bm\mu)] =
    \Tr(\bm\Sigma^{-1}\mathbf{M}) = \Tr(\mathbf{I}), \]
  since
  \[ \mathbf{M} = \Var[\mathbf{x} - \bm\mu] = \Var[\mathbf{x}] = \bm\Sigma. \]
  Thus, since we have that
  \[ \mathbb{E}\left[ -\frac{1}{2}\log |\bm\Sigma| \right] = -\frac{1}{2}\log
    |\bm\Sigma|, \]
  we get that
  \[ \mathbb{E}[\log f(\mathbf{x})] = -\frac{1}{2}\log |\bm\Sigma| + c \]
  for some $c \in \mathbb{R}$ that depends on $n$, but not on $\bm\mu$ or
  $\bm\Sigma$.
\end{proof}

In order to derive the ELBO, let us go back to \eqref{eq:elbo} and
write\footnote{At this point, we will drop the subscript denoting which
  variables the expectation is taken over. Also note that during the derivation
  equality is taken to mean ``equality up to an additive constant''.}
\[ \mathcal{L}(\bm\nu) = \mathbb{E}[\log\pfull] -
  \mathbb{E}[\log\approximation]. \]
By plugging in \eqref{full} and \eqref{eq:approximation}, we get
\[
  \begin{split}
    \mathcal{L}(\bm\nu) &= \mathbb{E}[\log p(\mathbf{X_u}) + \log p(\Theta |
    \mathbf{X_u}) + \log p(\mathbf{u} | \Theta, \mathbf{X_u}) + \log
    p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) + \log p(\mathcal{D} | r)] \\
    &- \mathbb{E}[\log q(\Theta) + \log q(\mathbf{u}) + \log q(\mathbf{r} |
    \Theta, \mathbf{u})].
  \end{split}
\]
Note that $\mathbb{E}[\log p(\mathbf{X_u})]$ is just a constant, so we can
simply drop it from the expression. Furthermore, since $q(\mathbf{r} | \Theta,
\mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u})$, they cancel
each other out. Then we can substitute various terms with their definitions to
get
\begin{alignat*}{3}
  \mathcal{L}(\bm\nu) &= \mathbb{E}\left[-\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1)\right] &&+ \mathbb{E}[\log
  \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu)] \\
  &+ \mathbb{E}\left[ \sum_{i=1}^N \sum_{t=1}^T Q_r(s_{i,t}, a_{i,t}) -
    V_r(s_{i,t}) \right] &&-
  \mathbb{E}\left[\log\left(\frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha -
        1}e^{-\beta\zeta}\right)\right] \\
  &- \mathbb{E}[\log\mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma)] &&-
  \mathbb{E}[\log\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})].
\end{alignat*}
Next, we can plug in the definitions of $\mathcal{N}(\mathbf{u}; \mathbf{0},
\Kuu)$ and $Q_r$, distribute the log across the gamma distribution, and apply
Lemma \ref{expectation_lemma} to the last two terms\footnote{We cannot
  apply the lemma to $\mathbb{E}[\log \mathcal{N}(\mathbf{u}; \mathbf{0},
  \Kuu)]$ because $\Kuu$ depends on $\Theta$.} to get
\begin{align*}
  \mathcal{L}(\bm\nu) &= \frac{1}{2}\log|\bm\Sigma| + \frac{1}{2}\log|\mathbf{S}| + \mathbb{E}\left[-\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1)\right] \\
  &+ \mathbb{E}\left[-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} -
    \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right] \\
  &+ \mathbb{E}\left[\sum_{i=1}^N \sum_{t=1}^T r(s_{i,t}) - V_r(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}}
    \mathcal{T}(s_{i,t}, a_{i,t}, s')V_r(s') \right] \\
  &- \mathbb{E}[\alpha\log\beta - \log\Gamma(\alpha) + (\alpha - 1)\log\zeta - \beta\zeta].
\end{align*}
Now we can remove $\mathbb{E}\left[-\frac{m}{2}\log2\pi\right]$ since it is
constant w.r.t. both the variational parameters and the variables the
expectation is over, and move constants (or variational parameters) independent
of the approximated variables outside of the expectations. Also note that
$\mathbb{E}[\zeta] = \alpha/\beta$ and
$\mathbb{E}[\log \zeta] = \psi(\alpha) - \log \beta$, where $\psi$ is the
digamma function defined as $\psi(x) =
\frac{d}{dx}\log\Gamma(x)$\cite{DBLP:books/lib/Bishop07}. This allows us to
simplify the expression to the following:
\begin{align*}
  \mathcal{L}(\bm\nu) &= \frac{1}{2}\log|\bm\Sigma| + \frac{1}{2}\log|\mathbf{S}| - \frac{1}{2} \mathbb{E}[\Tr[\Kuu^{-2}]] -
    \sum_{i=1}^d \mathbb{E}[\log(\lambda_i + 1)] \\
  &- \frac{1}{2}\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] - \frac{1}{2}\mathbb{E}[\log|\Kuu|] + \alpha - \log\beta + \log\Gamma(\alpha) + (1 - \alpha)\psi(\alpha) \\
  &+ \sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[r(s_{i,t})] - \mathbb{E}[V_r(s_{i,t})] + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')]. \\
\end{align*}

\subsection{Derivatives}

>>>>

\newcommand{\Kuu}{\mathbf{K}_{\mathbf{u},\mathbf{u}}}
\newcommand{\Krr}{\mathbf{K}_{\mathbf{r},\mathbf{r}}}
\newcommand{\Kru}{\mathbf{K}_{\mathbf{r},\mathbf{u}}}

\newcommand{\DKL}{D_{\mathrm{KL}}}

\newcommand{\pfull}{p(\mathcal{D}, \Theta, \mathbf{X_u}, \mathbf{u}, \mathbf{r})}
\newcommand{\pfullS}{p(\mathcal{D}, \Theta_s, \mathbf{X_u}, \mathbf{u}_s, \mathbf{r}_s)}
\newcommand{\posterior}{p(\Theta, \mathbf{u}, \mathbf{r} | \mathcal{D}, \mathbf{X_u})}
\newcommand{\approximation}{q_{\bm\nu}(\Theta, \mathbf{u}, \mathbf{r})}
\newcommand{\approximationS}{q_{\bm\nu}(\Theta_s, \mathbf{u}_s, \mathbf{r}_s)}

\newcommand{\Eq}{\mathbb{E}_{(\Theta, \mathbf{u}, \mathbf{r}) \sim \approximation}}

\newcommand{\dm}{\frac{\partial}{\partial\mathbf{m}}}
\newcommand{\dS}{\frac{\partial}{\partial\mathbf{S}}}
\newcommand{\da}{\frac{\partial}{\partial\alpha}}
\newcommand{\db}{\frac{\partial}{\partial\beta}}

\section{Introduction}
%briefly explain the context of the project problem
% TODO: introduce a data set
% TODO: example application for motivation (why would the world be better)
% TODO: (possibly) a figure with my notation
% TODO: explain the MDP Q/V stuff
% TODO: introduce research questions

@@

% TODO: positive definite
As training a GP with $n$ data points has a time complexity of
$\mathcal{O}(n^3)$ \cite{DBLP:books/lib/RasmussenW06}, numerous approximation
methods have been suggested, many of which select a subset of data called
\emph{inducing points} and focus most of the training effort on them
\cite{DBLP:journals/corr/abs-1807-01065}. Let $\mathbf{X_u}$ be the matrix of
features at inducing states, $\mathbf{u}$ the rewards at those states, and
$\mathbf{r}$ a vector with $r(\mathcal{S})$ as elements. Then the full joint
probability distribution can be factorised as
\begin{equation} \label{full}
  \pfull = p(\mathbf{X_u}) \times p(\Theta | \mathbf{X_u}) \times p(\mathbf{u}
  | \Theta, \mathbf{X_u}) \times p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}) \times p(\mathcal{D} | r).
\end{equation}
Here $p(\mathbf{X_u})$ and $p(\Theta | \mathbf{X_u})$ are customisable priors,
\begin{align}
  p(\mathbf{u} | \Theta, \mathbf{X_u}) &= \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu) \nonumber \\
                                       &= \frac{1}{(2\pi)^{m/2}|\Kuu|^{1/2}}\exp \left( -\frac{1}{2} \mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} \right) \nonumber \\
                                       &= \exp\left(-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} - \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right) \label{eq:normal}
\end{align}
is the GP prior \cite{DBLP:books/lib/RasmussenW06}, where $m \in \mathbb{N}$ is
the number of inducing points. The GP posterior is a
multivariate Gaussian \cite{DBLP:conf/nips/LevinePK11}
\begin{equation} \label{eq:r}
  p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) =
  \mathcal{N}(\mathbf{r}; \Kru^\intercal\Kuu^{-1}\mathbf{u}, \Krr - \Kru^\intercal\Kuu^{-1}\Kru),
\end{equation}
and $p(\mathcal{D} | r)$ is as in \eqref{pDr}. The matrices such as
$\Kru$ are called \emph{covariance matrices} and are defined as
$[\Kru]_{i,j} = k_\Theta(\mathbf{x}_{\mathbf{r},i}, \mathbf{x}_{\mathbf{u},j})$,
where $\mathbf{x}_{\mathbf{r},i}$ and $\mathbf{x}_{\mathbf{u},j}$ denote feature
vectors for the $i$th state in $\mathcal{S}$ and the $j$th state in
$\mathbf{X_u}$, respectively \cite{DBLP:conf/uai/JinDAS17}.

% TODO: where does the lower bound come from?
Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\Theta$, inducing rewards $\mathbf{u}$, and the reward function $r$. While the
previous paper that considered this IRL model computed maximum likelihood estimates
for $\Theta$ and $\mathbf{u}$, and made an assumption that $\mathbf{r}$ in
\eqref{eq:r} has zero variance \cite{DBLP:conf/nips/LevinePK11}, we aim to
avoid this assumption and use variational inference to approximate the full
posterior distribution $\posterior$. \emph{Variational inference} (VI) is an
approximation technique for probability densities \cite{blei2017variational}.
Let $\approximation$ be our approximating family of probability distributions
for $\posterior$ with its own hyperparameter vector $\bm\nu$. Then it is up
to VI algorithms to optimise $\bm\nu$ in order to minimise the
\emph{Kullback-Leibler} (KL) divergence between the original probability
distribution and our approximation.  KL divergence (asymmetrically) measures how
different the two distributions are and in this case can be defined as follows
\cite{blei2017variational}:
\[ \DKL(\approximation || \posterior) = \Eq \left[ \log
  \frac{\approximation}{\posterior} \right]. \]
Since KL divergence is typically hard to compute, instead of minimising it, VI
typically tries to maximise the \emph{evidence lower bound} (ELBO) defined as
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L}(\bm\nu) &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\Theta\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}

@@

% TODO: review IRL papers, their weaknesses

@@

At this point we are forced to make assumptions about the approximate posterior
in order to arrive at an implementable solution. Time permitting, ways to relax
the assumptions may be investigated towards the end of the project.

First, as is common in the literature for applying VI to GPs
\cite{DBLP:conf/nips/ChengB17,DBLP:conf/uai/HensmanFL13,DBLP:conf/icml/HoangHL15,DBLP:journals/jmlr/Titsias09},
we simply set
\begin{equation}
  q(\mathbf{r} | \Theta, \mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u},
  \mathbf{u}).
\end{equation}
We can make a similarly justified choice for $q(\mathbf{u} | \Theta)$:
\begin{equation}
  q(\mathbf{u} | \Theta) = q(\mathbf{u}) = \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})
\end{equation}
for some variational parameters $\mathbf{m} \in \mathbb{R}^{m}$ and a positive
semi-definite matrix $\mathbf{S} \in \mathbb{R}^{m \times m}$.
\cite{DBLP:conf/nips/ChengB17,DBLP:journals/jmlr/HensmanDS17,DBLP:conf/aaai/HoangHL17}.

% TODO: uncorrelated gammas instead of a multivariate Gaussian? review several options
In order to have a reasonable way of calculating/approximating the ELBO (see
Section \ref{vi_algs}), we need to have an approximating distribution for
$\Theta$, but unfortunately all the papers in Table \ref{table:approximations}
either fix it or treat it as a variational parameter. Hence we make our first
assumption without justification from previous literature:
\begin{equation} \label{eq:theta}
  q(\Theta) = q(\zeta) \times q(\bm\lambda).
\end{equation}
As it is reasonable for $\bm\lambda$ to take any value in $\mathbb{R}^d$, we can
approximate it simply as
\begin{equation} \label{eq:lambda}
  q(\bm\lambda) = \mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma)
\end{equation}
for some variational parameters $\bm\mu \in \mathbb{R}^d$ and $\bm\Sigma \in
\mathbb{R}^{d \times d}$. In contrast, as the covariance function must produce
non-negative values, we want to restrict $\zeta$ to positive real numbers. At
the same time, the distribution for $\zeta$ should have a flexible mean (i.e.,
not be tied to zero, like in the exponential distribution) and tails that
converge to zero as the value of the random variable moves away from the mean.
We might want to support some right skew, but the distribution should be close
to symmetric with at least some parameter values. This limits our choice of
distributions quite significantly, and we decide to go with the gamma
distribution as it is fairly flexible and commonly used
\cite{hogg2018introduction}. We then define the probability density function of
the gamma distribution as
\begin{equation} \label{eq:zeta}
  q(\zeta) = \Gamma(\zeta; \alpha, \beta) =
  \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta},
\end{equation}
where $\alpha > 0$ and $\beta > 0$ are two parameters of the distribution, and
$\Gamma(\cdot)$ is the gamma function. This gives us our vector of variational
parameters $\bm\nu = (\mathbf{m}, \mathbf{S}, \bm\mu, \bm\Sigma, \alpha,
\beta)^\intercal$.

\subsection{Evidence Lower Bound}
% TODO: remove/add niu from/to q everywhere

In this section we derive and simplify the ELBO for this (now fully specified)
model. In order to derive the ELBO, let us go back to \eqref{eq:elbo} and
write\footnote{At this point, we will drop the subscript denoting which
  variables the expectation is taken over. Also note that throughout the derivation
  equality is taken to mean ``equality up to an additive constant''.}
\[ \mathcal{L}(\bm\nu) = \mathbb{E}[\log\pfull] -
  \mathbb{E}[\log\approximation]. \]
By plugging in \eqref{full} and \eqref{eq:approximation}, we get
\[
  \begin{split}
    \mathcal{L}(\bm\nu) &= \mathbb{E}[\log p(\mathbf{X_u}) + \log p(\Theta |
    \mathbf{X_u}) + \log p(\mathbf{u} | \Theta, \mathbf{X_u}) + \log
    p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u}) + \log p(\mathcal{D} | r)] \\
    &- \mathbb{E}[\log q(\Theta) + \log q(\mathbf{u}) + \log q(\mathbf{r} |
    \Theta, \mathbf{u})].
  \end{split}
\]
Note that $\mathbb{E}[\log p(\mathbf{X_u})]$ is just a constant, so we can
simply drop it from the expression. Furthermore, since $q(\mathbf{r} | \Theta,
\mathbf{u}) = p(\mathbf{r} | \Theta, \mathbf{X_u}, \mathbf{u})$, they cancel
each other out. Then we can substitute various terms with their definitions to
get
\begin{alignat*}{3}
  \mathcal{L}(\bm\nu) &= \mathbb{E}\left[-\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1)\right] &&+ \mathbb{E}[\log
  \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu)] \\
  &+ \mathbb{E}\left[ \sum_{i=1}^N \sum_{t=1}^T Q_r(s_{i,t}, a_{i,t}) -
    V_r(s_{i,t}) \right] &&-
  \mathbb{E}\left[\log\left(\frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha -
        1}e^{-\beta\zeta}\right)\right] \\
  &- \mathbb{E}[\log\mathcal{N}(\bm\lambda; \bm\mu, \bm\Sigma)] &&-
  \mathbb{E}[\log\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})].
\end{alignat*}
For the last two terms we can use the fact that, e.g., $-\mathbb{E}[\log
\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})] = \frac{1}{2}\log|\mathbf{S}|$
up to an additive constant that depends on the number of dimensions of the
distribution \cite{DBLP:journals/tit/AhmedG89}. Note that we cannot use this
fact for $\mathbb{E}[\log \mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu)]$  because
$\Kuu$ depends on $\Theta$. We also plug in the definitions of
$\mathcal{N}(\mathbf{u}; \mathbf{0}, \Kuu)$ and $Q_r$, and get:
\begin{align*}
  \mathcal{L}(\bm\nu) &= \frac{1}{2}\log|\bm\Sigma| + \frac{1}{2}\log|\mathbf{S}| + \mathbb{E}\left[-\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1)\right] \\
  &+ \mathbb{E}\left[-\frac{1}{2}\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u} -
    \frac{1}{2}\log|\Kuu| - \frac{m}{2}\log 2\pi\right] \\
  &+ \mathbb{E}\left[\sum_{i=1}^N \sum_{t=1}^T r(s_{i,t}) - V_r(s_{i,t}) + \gamma\sum_{s' \in \mathcal{S}}
    \mathcal{T}(s_{i,t}, a_{i,t}, s')V_r(s') \right] \\
  &- \mathbb{E}[\alpha\log\beta - \log\Gamma(\alpha) + (\alpha - 1)\log\zeta - \beta\zeta].
\end{align*}
Now we can remove $\mathbb{E}\left[-\frac{m}{2}\log2\pi\right]$ since it is
constant w.r.t. both the variational parameters and the variables the
expectation is over, and move constants (or variational parameters) independent
of the approximated variables outside of the expectations. Also note that
$\mathbb{E}[\zeta] = \alpha/\beta$ and
$\mathbb{E}[\log \zeta] = \psi(\alpha) - \log \beta$, where $\psi$ is the
digamma function defined as $\psi(x) = \frac{d}{dx}\log\Gamma(x)$
\cite{DBLP:books/lib/Bishop07}. This allows us to
simplify the expression to the following:
\begin{align*}
  \mathcal{L}(\bm\nu) &= \frac{1}{2}\log|\bm\Sigma| + \frac{1}{2}\log|\mathbf{S}| - \frac{1}{2} \mathbb{E}[\Tr[\Kuu^{-2}]] -
    \sum_{i=1}^d \mathbb{E}[\log(\lambda_i + 1)] \\
  &- \frac{1}{2}\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] - \frac{1}{2}\mathbb{E}[\log|\Kuu|] + \alpha - \log\beta + \log\Gamma(\alpha) + (1 - \alpha)\psi(\alpha) \\
  &+ \sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[r(s_{i,t})] - \mathbb{E}[V_r(s_{i,t})] + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')]. \\
\end{align*}

\subsection{Derivatives}
% TODO: integrals always mean "on all range"
% TODO: some kind of intro

\subsubsection{$\partial/\partial\mathbf{m}$}
